---
title: Population Dynamics
author:
- M. Larsen^[Indiana University, Bloomington, IN]
abstract:
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    number_sections: no
geometry: margin=1in
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
rm(list=ls())

# Load packages and dependencies
#require(aggregate)
library(reshape); library(nlme); library(vegan)
library(igraph); library(bipartite)
library(plotrix);library(pander)
library(ggplot2); library(gridExtra)
library(nlme)
library(synchrony); library(astsa)
library(tidyverse)
library(lemon)
library(pastecs)
library(tseries)

# load dependencies
source("./bin/microscope.counts.R")
```

```{r global-opts, include = FALSE}
#knitr options
knitr::opts_chunk$set(fig.width=6.5, fig.height=3,
                      echo=FALSE, warning=FALSE, message=FALSE)
#ggplot options
theme_std <- function (base_size = 10, base_family = "") {
  theme_grey(base_size = base_size, base_family = base_family) %+replace% 
    theme(axis.ticks = element_line(colour = "black", size = 0.5), 
          legend.key = element_rect(colour = NA), 
          legend.background = element_rect(fill =NA, colour = NA),
          panel.background = element_rect(fill = "white", colour = NA), 
          panel.border = element_rect(colour = "black", fill=NA, size = 1),
          axis.line = element_line(colour = "black", size = 1),
          panel.grid.major = element_line(NA), 
          panel.grid.minor = element_line(NA), 
          axis.text  = element_text(size=rel(0.9), colour = "grey40"),
          #          axis.title  = element_text(size=rel(1.0)),
          axis.title.x = element_text(margin = unit(c(3, 0, 0, 0), "mm"),
                                      size=rel(1.25), colour = "grey30"),
          axis.title.y = element_text(margin = unit(c(0, 3, 0, 0), "mm"),
                                      size=rel(1.25), angle = 90, colour= "grey30"),
          strip.text = element_text(size = rel(1.25), colour = "black", face = "bold"), #face = "normal"
          strip.background = element_blank(),
          plot.margin=unit(c(10,10,10,10),"pt"), 
          panel.spacing = unit(25,"pt")
    )
}
```

<Data Reduction>

```{r, results = 'hide'}
# Reduce raw microscope field counts (ts-counts_raw) using the function bin\microscope.counts to create average cyanobacteria and phage counts per sampling date for each chemostat in the experiment
# microscope.counts("./data/ts-counts_raw.csv",  "ts-counts")
# Read in the processed data
ts.counts <- read.csv(file="./data/ts-counts.csv",  header = F)
```

```{r cid-cals}
# Creates long form data file for time series
#remove header row
cstat <- ts.counts[2:nrow(ts.counts),]
#remove factors from data set
cstat[,1] <- factor(cstat[,1])
cstat[,2] <- factor(cstat[,2])
cstat[,3] <- factor(cstat[,3])
cstat[,4] <- factor(cstat[,4])
cstat[,5] <- factor(cstat[,5])
#assign ids and day
cstat.ids <- ts.counts[1,1:5]
day <- as.numeric(ts.counts[1,6:ncol(ts.counts)])
#Generate information for each chemostat into long form
cID <- unique(cstat[,2])
abd.list <- c(); day.list <- c(); cID.list <- c(); type.list <- c()
lim.list <- c(); virus.list <- c()
for(i in 1:length(cID)){
  org <- unique(cstat[cstat[,2] == cID[i],1])
  
  for(j in 1:length(org)){
    ids <- cstat[(cstat[,2] == cID[i] & cstat[,1] == org[j]),1:5]
    tmp <- as.numeric(cstat[(cstat[,2] == cID[i] & cstat[,1] == org[j]),6:ncol(cstat)])
    tmpday <- day
    
    abd.list <- append(abd.list,tmp)
    day.list <- append(day.list,tmpday)
    lim.list <- append(lim.list,rep(as.character(ids[1,5]),length(tmpday)))
    virus.list <- append(virus.list,rep(as.character(ids[1,4]),length(tmpday)))
    cID.list <- append(cID.list,rep(as.character(cID[i]),length(day)))
    type.list <- append(type.list,rep(as.character(org[j]),length(day)))
  }
}
#merge data and change column names
cstat.abd <- data.frame(day.list,type.list,lim.list,virus.list,cID.list,abd.list)
colnames(cstat.abd) <- c("day","microbe","lim","type","cID","abd")
#write.csv(cstat.abd, file = "./cid-means.csv", row.names = F)
```


```{r trt-cals}
#reshape data file from long form to columns for syn, phage, and p:s ratio
cstat.abd <- read.csv("./data/cid-means.csv", header = T)
cstat.reshape <- cstat.abd
day.num <- day
cstat.reshape$day <- as.factor(cstat.reshape$day)
m1 <- melt(cstat.reshape, id = c("day","microbe","lim","type", "cID"))
c1 <- cast(m1, day +lim + type + cID ~ microbe)

#sort data and calculate p:s ratio at same time point
cstat.reshape <- c1[order(c1$type, c1$cID, c1$day),]
cstat.reshape$pst1 <- round(cstat.reshape[,5]/cstat.reshape[,6], digits = 2)
cstat.reshape$day.num <- rep(day.num,10)

#remove control data and truncate syn/phage data between days 0 and 166
cstat.infect <- subset(cstat.reshape,cstat.reshape$type == "Infect")
cstat.infect <- cstat.infect[,-3]
cstat.truc <- subset(cstat.infect,cstat.infect$day.num > 0, na.rm = T)

#write.csv(cstat.truc, file = "./output/cstatabd-stats.csv", row.names = F)

#Create mean and sem data calculations
sem <- function(x){
  sd(x)/sqrt(3)
}
c2 <- cast(m1, day + lim + type ~ microbe, mean)
cstat.means <- c2[order(c2$lim, c2$type, c2$day),]
m2 <- melt(cstat.means, id = c("day", "lim", "type"))
c3 <- cast(m2, day + lim ~ microbe + type)
cstat.means <- c3[order(c3$lim, c3$day),]
c4 <- cast(m1, day + lim + type ~ microbe, sem)
m3 <- melt(c4, id = c("day", "lim", "type"))
c5 <- cast(m3, day + lim ~ microbe + type)
cstat.sem <- c5[order(c5$lim, c5$day),]
avg.data <- merge(cstat.means,cstat.sem, c("day","lim"))
avg.data <- avg.data[order(avg.data$lim,avg.data$day),]
avg.data <- cbind(rep(day,2),avg.data[,2:ncol(avg.data)])
colnames(avg.data) <- c("day","lim",
                        "PmeanC","PmeanI","SmeanC","SmeanI",
                        "PseC","PseI","SseC","SseI")
#write.csv(avg.data, file = "./data/cstat-means.csv", row.names = F)
```

## Figure 1: Average population dynamics

```{r avgpops-revised,results = 'hide', echo=FALSE, fig.asp='75%'}
avg <- read.csv("./data/cstat-means.csv", header = T)
day.start <- -6
day.end <- 170

# Used for saving pub quality figures
# tiff(filename = "./supporting-files/Figure1_AvgPopDynamics.tiff",
#     width = 1140, height = 570, units = "px", pointsize = 13,
#     bg = "white", res = 100)

par(mfrow = c(1,2))
par(mar = c(4,6,3,0)+0.1)
plot(avg$day[avg$day >= day.start],avg$NmeanSI[avg$day >= day.start],
     xlim = c(day.start,day.end), ylim = c(1*10^5,10^9),
     main = "N-limited",
     log = "y", xaxt = 'n',yaxt = 'n',
     xlab = expression(bold("Time (d)")), 
     ylab = expression(bold(paste("Abundance (mL"^"-1",")"))),
     cex.lab = 1.5, font.lab = 2,cex.main = 1.5,
     lty = 1, lwd = 2, type = "o", col = "black", bg = "white", 
     pch = 21, cex = 1)
if(day.start <= 0){
  abline(v = 0, lwd = 2, lty = 1, col = "black")
  rect(day.start-12,1*10^3,0,10^10, col = "grey94")
}
if(day.start > -6){
  text(-60,10^5,labels = "Acclimation", font = 2, cex = 1.15)
  text(80,10^5,labels = "Phage addition", font = 2, cex = 1.15)
}

axis(1, cex.axis = 1.15,c(-120,-80,-40,0,40,80,120,160))
ticks <- seq(5, 9, by=1)
labels <- sapply(ticks, function(i) as.expression(bquote(10^ .(i))))
axis(2, cex.axis = 1.15, at = c(10^5, 10^6, 10^7, 10^8, 10^9), labels = labels, las = 1)
#abline(v = 0, col = "grey", lty = 2, lwd = 2)
box(lwd=3)

arrows(avg$day[avg$day >= day.start],
       avg$NmeanSI[avg$day >= day.start]-avg$NseSI[avg$day >= day.start],
       avg$day[avg$day >= day.start],
       avg$NmeanSI[avg$day >= day.start]+avg$NseSI[avg$day >= day.start],
       code=0,lwd=2.25)
arrows(avg$day,avg$NmeanP-avg$NseP,avg$day,avg$NmeanP+avg$NseP,code=0,lwd=2.25)

points(avg$day[avg$day >= day.start],avg$NmeanSI[avg$day >= day.start],
       pch = 21, col = "black",bg = "white",
       type = 'o',lwd = 2,cex = 1.25, lty = 1)
points(avg$day,avg$NmeanP,
       pch = 22, col = "black",bg = "white",
       type = 'l',lwd = 2,cex = 1.25 , lty = 1)
points(avg$day,avg$NmeanP,
       pch = 22, col = "grey50",bg = "black",
       type = 'p',lwd = 1.25,cex = 1.25 , lty = 1)
#text("a",cex=2.5,x=162,y=10^8.8)
par(mar = c(4,4,3,2)+0.1)
#PL
plot(avg$day[avg$day >= day.start],avg$PmeanSI[avg$day >= day.start],
     xlim = c(day.start,day.end),ylim = c(1*10^5,10^9), 
     main = "P-limited",
     log = "y", xaxt = 'n',yaxt = 'n',
     xlab = expression(bold("Time (d)")), ylab = NA,cex.lab = 1.5, cex.main = 1.5,
     lty = 1, lwd = 2, type = "o", col = "black", bg = "white", 
     pch = 21, cex = 1, font.lab = 2)
if(day.start <= 0){
  abline(v = 0, lwd = 2, lty = 1, col = "black")
  rect(day.start-12,1*10^3,0,10^10, col = "grey94")
}
if(day.start > -6){
  text(-60,10^4,labels = "Acclimation", font = 2, cex = 1.2)
  text(80,10^4,labels = "Phage addition", font = 2, cex = 1.2)
}
axis(1, cex.axis = 1.15,c(-120,-80,-40,0,40,80,120,160))
ticks <- seq(4, 9, by=1)
labels <- sapply(ticks, function(i) as.expression(bquote(10^ .(i))))
axis(2, cex.axis = 1.15, at = c(10^4,10^5, 10^6, 10^7, 10^8, 10^9), labels = labels, las = 1, font = 2)
box(lwd=3)

arrows(avg$day[avg$day >= day.start],
       avg$PmeanSI[avg$day >= day.start]-avg$PseSI[avg$day >= day.start],
       avg$day[avg$day >= day.start],
       avg$PmeanSI[avg$day >= day.start]+avg$PseSI[avg$day >= day.start],code=0,lwd=2.25)
arrows(avg$day,avg$PmeanP-avg$PseP,avg$day,avg$PmeanP+avg$PseP,code=0,lwd=2.25)
points(avg$day[avg$day >= day.start],avg$PmeanSI[avg$day >= day.start],
       pch = 21, col = "black",bg = "white",
       type = 'o',lwd = 2,cex = 1.25, lty = 1)
points(avg$day,avg$PmeanP,
       pch = 22, col = "black",bg = "black",
       type = 'l',lwd = 2,cex = 1.25 , lty = 1)
points(avg$day,avg$PmeanP,
       pch = 22, col = "grey50",bg = "black",
       type = 'p',lwd = 1.25,cex = 1.25 , lty = 1)
#text("b",cex=2.5,x=162,y=10^8.8)
legend("bottomright", c("phage","cyanobacteria"),
       pch = c(22,21), col = "black", 
       pt.bg = c("black","white"),pt.cex = 1.25, pt.lwd = 2,
       text.font = c(1,1),
       bty = 'n', cex = 1, lwd = 1)
#dev.off()
```

## Figure S2: Average population dynamics

```{r avgpops-ctrl_revised}
#tiff(filename = "../../figures/FigS1-CtrlAvgPopDynamics.tiff",
#     width = 1140, height = 570, units = "px", pointsize = 13,
#     bg = "white", res = 100)
par(mar = c(5,6,2,0)+0.1, mfrow = c(1,2))
plot(avg$day[avg$day >= day.start],avg$NmeanC[avg$day >= day.start],
     xlim = c(day.start,day.end),ylim = c(1*10^5,10^9), log = "y", xaxt = 'n',yaxt = 'n',
     xlab = "Time (d)", ylab = expression(bold(paste("Abundance (mL"^"-1",")"))),
     main = "N-limited",
     cex.lab = 1.5, font.lab = 2, cex.main = 1.5,
     lty = 1, lwd = 2, type = "o", col = "black", bg = "grey",
     pch = 21, cex = 1)

axis(1, cex.axis = 1.15,c(-120,-80,-40,0,40,80,120,160))
ticks <- seq(4, 9, by = 1)
labels <- sapply(ticks, function(i) as.expression(bquote(10^ .(i))))
axis(2, cex.axis = 1.15, at = c(10^4,10^5, 10^6, 10^7, 10^8, 10^9), 
     labels = labels, las = 1)

if(day.start <= 0){
  abline(v = 0, lwd = 2, lty = 1, col = "black")
  rect(day.start-12,1*10^3,0,10^10, col = "grey94")
}
box(lwd=3)

arrows(avg$day[avg$day >= day.start],
       avg$NmeanC[avg$day >= day.start]-avg$NseC[avg$day >= day.start],
       avg$day[avg$day >= day.start],
       avg$NmeanC[avg$day >= day.start]+avg$NseC[avg$day >= day.start],code=0,lwd=1.5)
points(avg$day[avg$day >= day.start],avg$NmeanC[avg$day >= day.start],
       pch = 21, col = "black",bg = "grey",
       type = 'o',lwd = 1.5,cex = 1.5, lty = 1)
par(mar = c(5,4,2,2)+0.1)
plot(avg$day[avg$day >= day.start],avg$PmeanC[avg$day >= day.start],
     xlim = c(day.start,day.end),ylim = c(1*10^5,10^9), log = "y", xaxt = 'n',yaxt = 'n',
     xlab = "Time (d)", ylab = NA, 
     main = "P-limited",
     cex.main = 1.5, cex.lab = 1.5, 
     lty = 1, lwd = 2, type = "o", col = "black", bg = "grey", 
     pch = 21, cex = 1, font.lab = 2)

axis(1, cex.axis = 1.15,c(-120,-80,-40,0,40,80,120,160))
ticks <- seq(4, 9, by=1)
labels <- sapply(ticks, function(i) as.expression(bquote(10^ .(i))))
axis(2, cex.axis = 1.15, at = c(10^4,10^5, 10^6, 10^7, 10^8, 10^9), 
     labels = labels, las = 1, font = 2)
if(day.start <= 0){
  abline(v = 0, lwd = 2, lty = 1, col = "black")
  rect(day.start-12,1*10^3,0,10^10, col = "grey94")
}
box(lwd=3)

arrows(avg$day[avg$day >= day.start],
       avg$PmeanC[avg$day >= day.start]-avg$PseC[avg$day >= day.start],
       avg$day[avg$day >= day.start],
       avg$PmeanC[avg$day >= day.start]+avg$PseC[avg$day >= day.start],code=0,lwd=1.5)
points(avg$day[avg$day >= day.start],avg$PmeanC[avg$day >= day.start],
       pch = 21, col = "black",bg = "grey",
       type = 'o',lwd = 1.5,cex = 1.5, lty = 1)
#text("b",cex=2.5,x=162,y=10^8.8)
#dev.off()
```

\newpage

## Figure 1: Average population dynamics - revised

```{r cstatSummaryCalcs}
sem <- function(x){
  sd(x)/sqrt(3)
}

ts.counts.mod <- ts.counts
colnames(ts.counts.mod) <- c("microbe","cID","rep","trt","lim",ts.counts[1,6:ncol(ts.counts)])

ts.counts.mod <- ts.counts.mod[-1,]

ts.counts.sum <- ts.counts.mod %>%
  gather(day, abundance, "-125":"166") %>%
  mutate(day = as.numeric(day), 
         trt.lim = paste(trt,".",lim, sep =""),
         microbe.cID = paste(microbe,".",cID, sep ="")) %>%
  group_by(day, trt, lim, microbe) %>%
  summarise(mean = mean(abundance, na.rm = T), sem = sem(abundance), 
            min = min(abundance), max = max(abundance))
```

```{r treatmentdynamics-figure}
#droplevels(ts.counts.sum$lim)
#levels(ts.counts.sum$lim) <- c("","N-limited","P-limited")

# generate plot
p <- ggplot(data = ts.counts.sum %>% 
              filter(trt == "Infect", day >= -6), 
            aes(x = day, y = mean, ymin = mean-sem, ymax = mean+sem,
                shape = microbe, col = microbe)) +
  geom_vline(xintercept = 0, size = 0.25) +
  geom_errorbar(col = "black", width = 0)+
  geom_line(col = "black") + geom_point() + facet_grid(~lim) +
  scale_shape_manual(values = c(15,19), name = "") + 
  geom_point(data = ts.counts.sum %>% 
               filter(trt == "Infect", microbe =="Syn", day >= -6), 
             aes(x = day, y = mean), size = 2, col= "black") +
  geom_point(data = ts.counts.sum %>% 
               filter(trt == "Infect", microbe =="Syn", day >= -6), 
             aes(x = day, y = mean), size = 1.5, col= "white") +
  scale_color_manual(values = c("black","white"), name = "")+
  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Time (d)", y = expression(Abundance~(mL^-1))) + 
  theme_std() +
  annotation_logticks(sides = "lr")

# move legend
reposition_legend(p, panel = "panel-2-1",'bottom right')

# save plot
#ggsave(plot = reposition_legend(p, panel = "panel-2-1",'bottom right'), filename = "../../figures/Fig1_AvgPopDynamics_revised.tiff", units = "in", height = 3, width = 5.75, dpi = 300, scale = 1.5, device = "tiff")
```

## Figure S2: Average population dynamics - revised

```{r controldynamics-figure}
p <- ggplot(data = ts.counts.sum %>% 
              filter(trt == "Control", day >= -6), 
            aes(x = day, y = mean, ymin = min, ymax = max)) +
  geom_vline(xintercept = 0, size = 0.25) +
  geom_errorbar(col = "black", width = 0)+
  geom_line(col = "grey70") + geom_point() + facet_grid(~lim) +
  #scale_shape_manual(values = c(15,19)) + 
  scale_color_manual(values = c("black","white"))+
  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Time (d)", y = expression(Abundance~(mL^-1))) + 
  theme_std() +
  annotation_logticks(sides = "lr")

print(p)

# ggsave(plot = p, filename = "./supporting-files/FigureS2_AvgPopDynamics_revised.tiff", units = "in", height = 3, width = 5.75, dpi = 300, scale = 1.2, device = "tiff")
```

\newpage

<CCF analysis>
```{r Nlim_CCF, eval = F}
# convert irregular time series to regular time series

ts.counts.sum2 <- as.data.frame(ts.counts.sum)
ccf.dataset.N.log <- ts.counts.sum2 %>%
  filter(day > 0, day<= 149, trt == "Infect", lim == "N-limited") %>%
  select(microbe, day, mean) %>%
  ungroup() %>%
  mutate(mean = log10(mean)) %>%
  spread(microbe, mean)
length(ccf.dataset.N.log$day)
intervals.N <- ccf.dataset.N.log$day[2:59] - ccf.dataset.N.log$day[1:58]
range(intervals.N); mean(intervals.N)

# range is variable - need to be regulated to be converted to ts
ts.n.reg <- regul(ccf.dataset.N.log$day, ccf.dataset.N.log[2:3], xmin = 2, deltat = 2,
                  methods = c("l","l"), n = 74)
#tiff(filename = "../../figures/SRegTS_N.tiff", units = "in", width = 7.5, height = 4, res = 300)
par(mfrow = c(1,2))
n.syn <- plot(ts.n.reg,2)
n.phage <- plot(ts.n.reg,1)
#dev.off()
specs(ts.n.reg)
syn.n.ts <- extract(ts.n.reg, series = "Syn")
phage.n.ts <- extract(ts.n.reg, series = "Phage")

adf.test(syn.n.ts)
diff.syn.n.ts <- diff(syn.n.ts)
adf.test(diff.syn.n.ts)
adf.test(phage.n.ts)
diff.phage.n.ts <- diff(phage.n.ts)
adf.test(diff.phage.n.ts)
#ccf.n <- ccf(syn.n.ts,phage.n.ts, type = "correlation")
ccf.n <- ccf(diff.syn.n.ts,diff.phage.n.ts, type = "correlation", main ="N-limited", lag.max = 5)
alldata=ts.intersect(diff.syn.n.ts,synlag2=stats::lag(diff.syn.n.ts,2), 
                     diff.syn.n.ts,synlag4=stats::lag(diff.syn.n.ts,4),
                     diff.syn.n.ts,synlag6=stats::lag(diff.syn.n.ts,6))
str(alldata) #ts from 4 to 136
alldata <- as.data.frame(alldata)
tryit = lm(diff.phage.n.ts[2:68]~synlag2+synlag4 +synlag6, data = alldata)
summary (tryit)
acf2(residuals(tryit))
```

```{r Plim_CCF, eval = F}
# convert irregular time series to regular time series
ccf.dataset.P.log <- ts.counts.sum2 %>%
  filter(day > 0, day<= 149, trt == "Infect", lim == "P-limited") %>%
  select(microbe, day, mean) %>%
  ungroup() %>%
  mutate(mean = log10(mean)) %>%
  spread(microbe, mean)
length(ccf.dataset.P.log$day)
intervals.P <- ccf.dataset.P.log$day[2:59] - ccf.dataset.P.log$day[1:58]
range(intervals.P); mean(intervals.P)
# range is variable - need to be regulated to be converted to ts
ts.p.reg <- regul(ccf.dataset.P.log$day, ccf.dataset.P.log[2:3], xmin = 2, deltat = 2,
                  methods = c("l","l"), n = 74)
tiff(filename = "../../figures/SRegTS_P.tiff", units = "in", width = 7.5, height = 4, res = 300)
par(mfrow = c(1,2))
p.syn <- plot(ts.p.reg,2)
p.phage <- plot(ts.p.reg,1)
dev.off()
specs(ts.p.reg)
syn.p.ts <- extract(ts.p.reg, series = "Syn")
phage.p.ts <- extract(ts.p.reg, series = "Phage")
adf.test(syn.p.ts)
diff.syn.p.ts <- diff(syn.p.ts)
adf.test(diff.syn.p.ts)
adf.test(phage.p.ts)
diff.phage.p.ts <- diff(phage.p.ts)
adf.test(diff.phage.p.ts)
#ccf.n <- ccf(syn.n.ts,phage.n.ts, type = "correlation")
ccf.p <- ccf(diff.syn.p.ts,diff.phage.p.ts, type = "correlation", main ="P-limited", lag.max = 5)
```

```{r CCF_plot, eval = F}
tiff(filename = "../../figures/S_ccf.tiff", units = "in", width = 7.5, height = 3.5, res = 300)
par(mfrow = c(1,2))
ccf2(diff.syn.n.ts,diff.phage.n.ts, main ="N-limited", max.lag = 5)
ccf2(diff.syn.p.ts,diff.phage.p.ts, main ="P-limited", max.lag = 5)
dev.off()
```


\newpage

## Calculations and Statistics
### Reapeated Measures ANOVA (RMANOVA)

```{r EDA-plot}
p <- ggplot(data = ts.counts.sum %>% 
              filter(microbe == "Syn", day >= -6), 
            aes(x = day, y = mean, ymin = min, ymax = max, 
                col = trt, shape = trt)) +
  geom_vline(xintercept = 0, size = 0.25) +
  geom_errorbar(col = "grey70", width = 0)+
  geom_line(col = "grey25") + geom_point() + facet_grid(~lim) +
  #scale_shape_manual(values = c(15,19)) + 
  scale_color_manual(values = c("black","grey"))+
  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  labs(x = "Time (d)", y = expression(Abundance~(mL^-1))) + 
  theme_std() +
  annotation_logticks(sides = "lr")

print(p)
```

```{r NvP-phage, echo=FALSE}
# this is a mixed equation model where day and treatment are fixed factors, 
# while cID is a random factor
ts.abd <- subset(cstat.abd,cstat.abd$type == "Infect" & cstat.abd$day > 0)
ts.abd <- na.omit(ts.abd)
ts.abd$log.abd <- log10(ts.abd$abd)                   #Create a column of log data
ts.abd$day.fac <- as.factor(ts.abd$day)               #Convert time into a factor
phage <- subset(ts.abd,ts.abd$microbe == "Phage")
syn <- subset(ts.abd,ts.abd$microbe == "Syn")

# Phage - model fitting
model.ar <- lme(log.abd ~ lim * day.fac, random = ~ 1 | cID, 
                 correlation=corAR1(form = ~ 1 | cID),
                 data = phage)

model.arma <- lme(log.abd ~ lim * day.fac, random = ~1 | cID,
                  correlation = corARMA(form = ~1 | cID, p = 1, q = 1),
                  data = phage)

model.arma1 <- lme(log.abd ~ lim * day.fac, random = ~1 | cID,
                   correlation = corARMA(form = ~1 | cID, p = 1, q = 2),
                   data = phage)

mod.comp = anova(model.ar, model.arma, model.arma1)

#xtable(anova.lme(ar1.phage, type = "marginal"))
#summary(ar1.phage)
set.caption("RMANOVA table for SRIM8 phage between N-limited vs. P-limited environments. lim: nutrient limitation (nitrogen-limited [NL] or phosphorus-limited [PL]); day.fac: time in the experimental timecourse")
pander(anova.lme(model.ar, type = "marginal"))
```

```{r NvP-syn}
#Syn
model.ar <- lme(log.abd ~ lim * day.fac, random = ~ 1 | cID, 
                 correlation=corAR1(form = ~ 1 | cID),
                 data = syn)

model.arma <- lme(log.abd ~ lim * day.fac, random = ~1 | cID,
                  correlation = corARMA(form = ~1 | cID, p = 1, q = 1),
                  data = syn)

model.arma1 <- lme(log.abd ~ lim * day.fac, random = ~1 | cID,
                   correlation = corARMA(form = ~1 | cID, p = 1, q = 2),
                   data = syn)

mod.comp = anova(model.ar, model.arma, model.arma1)

#anova.lme(ar1.syn, type = "marginal")
#summary(ar1.syn)
set.caption("RMANOVA table for +Ph Synechococcus between N-limited and P-limited environments")
pander(anova.lme(model.arma, type = "marginal"))
```

\newpage

```{r TreatmentvControl-NL}
ts.abd <- cstat.abd %>%
  filter(microbe == "Syn",
         lim == "N",
         day >0) %>%
  mutate(log.abd = log10(abd),
         day.fac = as.factor(day)) %>%
  na.omit()

model.ar <- lme(log.abd ~ day.fac * type, random = ~ 1 | cID, 
                 correlation=corAR1(form = ~ 1 | cID),
                 data = ts.abd)

model.arma <- lme(log.abd ~ day.fac * type, random = ~1 | cID,
                  correlation = corARMA(form = ~1 | cID, p = 1, q = 1),
                  data = ts.abd)

model.arma1 <- lme(log.abd ~ day.fac * type, random = ~1 | cID,
                   correlation = corARMA(form = ~1 | cID, p = 1, q = 2),
                   data = ts.abd)

mod.comp = anova(model.ar, model.arma, model.arma1)

set.caption("RMANOVA table comparison between -Ph and +Ph Synechococcus under N-limitation. day.fac: time in experimental timecourse; type: *Synechococcus* exposure to phage [control, treatment]")
pander(anova.lme(model.arma1, type = "marginal"))  

```

```{r TreatmentvControl-PL}
ts.abd <- cstat.abd %>%
  filter(microbe == "Syn",
         lim == "P",
         day >0) %>%
  mutate(log.abd = log10(abd),
         day.fac = as.factor(day)) %>%
  na.omit()

model.ar <- lme(log.abd ~ day.fac * type, random = ~ 1 | cID, 
                 correlation=corAR1(form = ~1 | cID),
                 data = ts.abd)

model.arma <- lme(log.abd ~ day.fac * type, random = ~1 | cID,
                  correlation = corARMA(form = ~1 | cID, p = 1, q = 1),
                  data = ts.abd)

model.arma1 <- lme(log.abd ~ day.fac * type, random = ~1 | cID,
                   correlation = corARMA(form = ~1 | cID, p = 1, q = 2),
                   data = ts.abd)

mod.comp = anova(model.ar, model.arma, model.arma1)

set.caption("RMANOVA table comparison between -Ph and +Ph Synechococcus under P-limitation. day.fac: time in experimental timecourse; type: *Synechococcus* exposure to phage [control, treatment]")
pander(anova.lme(model.arma, type = "marginal"))  

```

```{r NvP-Controls}
ts.abd <- cstat.abd %>%
  filter(microbe == "Syn",
         type == "Control",
         day >0) %>%
  mutate(log.abd = log10(abd),
         day.fac = as.factor(day)) %>%
  na.omit()

model.ar <- lme(log.abd ~ day.fac * lim, random = ~ 1 | cID, 
                 correlation=corAR1(form = ~1 | cID),
                 data = ts.abd)

model.arma <- lme(log.abd ~ day.fac * lim, random = ~1 | cID,
                  correlation = corARMA(form = ~1 | cID, p = 1, q = 1),
                  data = ts.abd)

model.arma1 <- lme(log.abd ~ day.fac * lim, random = ~1 | cID,
                   correlation = corARMA(form = ~1 | cID, p = 1, q = 2),
                   data = ts.abd)

mod.comp = anova(model.ar, model.arma, model.arma1)

set.caption("RMANOVA table comparison between N-limited and P-limited control (-Ph) Synechococcus. day.fac: time in experimental timecourse; type: *Synechococcus* exposure to phage [control, treatment]")
pander(anova.lme(model.ar, type = "marginal"))  

```

\newpage

### Stability (1/CV) and associated tables

```{r}

ts.stab <- function(x){
  1/(sd(x)/mean(x))
}

dat = subset(cstat.abd,cstat.abd$day > 0)
dat$logabd <- log10(dat$abd)
dat = na.omit(dat)

means = cast(dat, lim + type ~ microbe, mean, value = "abd")
sems = cast(dat, lim + type ~ microbe, sem, value = "abd")
ts.stabs = cast(dat, lim + type ~ microbe, ts.stab, value = "logabd")

# create table
table1 <- data.frame("Limitation" = means[,1], 
                     "Treatment" = means[,2],
                     "Synechococcus Mean (SEM)" = paste(prettyNum(means$Syn, digits = 2, format = "e"),"(",prettyNum(sems$Syn, digits = 1, format = "e"),")",sep = ""),
                     "Syn Stability" = prettyNum(ts.stabs$Syn, format = "e", digits = 2),
                     "Phage Mean (SEM)" = paste(prettyNum(means$Phage, digits = 2),"(",prettyNum(sems$Phage, digits = 2, format = "e"),")",sep = ""),
                     "Phage Stability" = prettyNum(ts.stabs$Phage, digits = 2, format = "e"))
colnames(table1) <- c("Limitation","Treatment","Synechococcus mean densitiy (+/- SEM)","Synechococcus mean stability","Phage mean density (+/- SEM)", "Phage mean stability")

# Creates table by treatment
pander(table1, split.cells = c(5,10,15,15,15,15),split.tables = Inf, caption = "Average population density and stability by treatment")
means.cid = cast(dat, cID + type ~ microbe, mean, value = "abd")
sems.cid = cast(dat, cID + type ~ microbe, sem, value = "abd")
ts.stabs.cid = cast(dat, cID + type ~ microbe, ts.stab, value = "logabd")

table2 <- data.frame("Limitation" = means.cid[,1], 
                     "Treatment" = means.cid[,2],
                     "Synechococcus Mean (SEM)" = paste(prettyNum(means.cid$Syn, digits = 2, format = "e"),"(",prettyNum(sems.cid$Syn, digits = 1, format = "e"),")",sep = ""),
                     "Syn Stability" = prettyNum(ts.stabs.cid$Syn, format = "e", digits = 2),
                     "Phage Mean (SEM)" = paste(prettyNum(means.cid$Phage, digits = 2),"(",prettyNum(sems.cid$Phage, digits = 2, format = "e"),")",sep = ""),
                     "Phage Stability" = prettyNum(ts.stabs.cid$Phage, digits = 2, format = "e"))
colnames(table2) <- c("Chemostat","Treatment","Synechococcus mean densitiy (+/- SEM)","Synechococcus mean stability","Phage mean density (+/- SEM)", "Phage mean stability")
pander(table2, split.cells = c(5,10,15,10,15,10),split.tables = Inf, caption = "Average population density and stability by chemostat")
```

\newpage

### Topolographic statistics and associated tables

```{r ts-mean}
ts.abd <- subset(cstat.abd,cstat.abd$type == "Infect" & cstat.abd$day > 0)
ts.abd <- na.omit(ts.abd)
#Create function for the standard error of the mean. 
#The density measurements are based on the average of 10 microscope images.
sem <- function(x){
  sd(x)/sqrt(3)
}
#Calculate basic statistics including mean, variance, sd, sem for each pop
basic.stats <- matrix(NA, 12, 6)                       #make results storage
basic.stats <- as.data.frame(basic.stats)
colnames(basic.stats) <- c("lim","cID","microbe",
                           "mean","var","sem")
r <- 0
for(i in unique(ts.abd$cID)){  
  for(j in unique(ts.abd$microbe)){
    #subset data
    tmp <- ts.abd[ts.abd$microbe == j & ts.abd$cID == i,]
    r <- r + 1
    
    basic.stats[r,1] <- as.character(tmp[1,3])          #Limitation
    basic.stats[r,2] <- as.character(i)                 #cID
    basic.stats[r,3] <- as.character(j)                 #microbe
    
    basic.stats[r,4] <- mean(tmp[,6])                   #mean
    basic.stats[r,5] <- var(tmp[,6])                    #variance
    basic.stats[r,6] <- sem(tmp[,6])                    #SEM
  }
}
basic.stats <- basic.stats[order(basic.stats$microbe),]
rownames(basic.stats)=NULL
#pander(basic.stats)
```

```{r ts-minmax}
ts.stab <- function(x){
  1/(sd(x)/mean(x))
}
ts.abd <- subset(cstat.abd,cstat.abd$type == "Infect" & cstat.abd$day > 0)
ts.abd$logabd <- log10(ts.abd$abd)
ts.abd <- na.omit(ts.abd)
#Initialize data storage
results <- matrix(NA,12,10)
results <- as.data.frame(results)
colnames(results) <- c("lim", "cID", "microbe", "stab", 
                       "start.abd","final.abd",
                       "min.day", "min.abd", "max.day","max.abd")
#,"dec.rate", "recov.rate", "recov.final")
r <- 0
for(i in unique(ts.abd$cID)){  
  for(j in unique(ts.abd$microbe)){
    #subset data
    
    if(j == "Phage"){
      tmp <- ts.abd[ts.abd$microbe == j & ts.abd$cID == i,1:7]
    } else {
      tmp <- ts.abd[ts.abd$microbe == j & ts.abd$cID == i,1:7]
    }
    
    r <- r + 1 
    
    #topology calculations
    results[r,1] <- as.character(tmp[1,3])    #Limitation
    results[r,2] <- as.character(i)           #cID
    results[r,3] <- as.character(j)           #microbe   
    results[r,4] <- ts.stab(tmp[,7])          #population stability
    
    results[r,5] <- tmp[1,6]# starting abundance
    results[r,6] <- tmp[nrow(tmp),6]# final abundance
    results[r,7] <- tmp[, 1][tmp[, 6] == min(tmp[, 6])]# minimum day
    results[r,8] <- min(tmp[, 6])# minimum abd
    results[r,9] <- tmp[, 1][tmp[, 6] == max(tmp[, 6])]# max day
    results[r,10] <- max(tmp[, 6])# max abd
    #results[i,11] <- (tmp[1,6]-min(tmp[, 6]))/(tmp[, 1][tmp[, 6] == min(tmp[, 6])]) # decline rate
    #results[i,13] <- (Max.abd-Min.abd)/(Max.day-Min.day)# recovery rate
    #results[i,14] <- (TF-Min.abd)/(TFday-Min.day)# recovery rate final
    
  }
}
ts.top <- results
ts.top <- merge(basic.stats,ts.top,id=c("lim","cID","microbe"))
top.sort <- ts.top[order(ts.top$microbe),]
rownames(top.sort)=NULL
#colnames(top.sort)=c("Limitation Treatment", "Chemostat ID","Microbe","Mean Abundance","Variation", )
#write.csv(top.sort, file = "../../data/pop-dynamics/R/ts-topographic-info.csv",row.names = FALSE)
set.caption("Descriptive statistical summary of population data.")
pander(top.sort)
```

```{r summary-table}
#subset the data set
pphage <- subset(top.sort, top.sort$lim == "P" & top.sort$microbe == "Phage")
nphage <- subset(top.sort, top.sort$lim == "N" & top.sort$microbe == "Phage")
psyn <- subset(top.sort, top.sort$lim == "P" & top.sort$microbe == "Syn")
nsyn <- subset(top.sort, top.sort$lim == "N" & top.sort$microbe == "Syn")

#calculate the stats
syn.mean <- t.test(psyn$mean,nsyn$mean, alternative="two.sided",conf.int = 0.95)
phage.mean <- t.test(pphage$mean,nphage$mean, alternative="two.sided",conf.int = 0.95)

syn.stab <- t.test(psyn$stab,nsyn$stab, alternative="greater",conf.int = 0.95)
phage.stab <- t.test(pphage$stab,nphage$stab, alternative="greater",conf.int = 0.95)

#assemble the data
top.stats <- matrix(NA,4,5)
top.stats <- data.frame(top.stats)
colnames(top.stats) <- c("microbe","analysis","t","df","pvalue")
top.stats[,1] <- c("syn","syn","phage","phage")
top.stats[,2] <- c("mean","stability","mean","stability")
top.stats[,3] <- round(c(syn.mean[[1]],syn.stab[[1]],phage.mean[[1]],phage.stab[[1]]), digits = 3)
top.stats[,4] <- round(c(syn.mean[[2]],syn.stab[[2]],phage.mean[[2]],phage.stab[[2]]),digits = 2)
top.stats[,5] <- round(c(syn.mean[[3]],syn.stab[[3]],phage.mean[[3]],phage.stab[[3]]), digits = 3)

#top.stats
set.caption("t-test summary for population mean and stability of +Ph Synechococcus and phage")
pander(top.stats)
```